### 2009-08-31 paper/2009-09/memory.tex
More work reformatting and reorganizing paper

\section{Parallel Clich\'{e}s\label{s:cliche}}

\subsection{Operations\label{s:cliche-ops}}

This list details some operations which are supported by many parallel programming systems.
The toys described in Section~\ref{s:toys} provide opportunities for exercising many of these,
and implementors are encouraged to phrase discussion of their work in terms of these operations where appropriate.

\begin{itemize}
\item	elementwise operations on arrays (unary, binary, and scalar promotion)
\item	cumulative operations on arrays (reduction and parallel prefix)
\item	array re-shaping and re-sizing (e.g.\ sub-array extraction)
\item	partial (conditionally masked) versions of all of the above
\item	regular data motion (e.g.\ circular and planar shifting)
\item	irregular data motion (e.g.\ 1-to-1, 1-to-many, and many-to-1 permutation)
\item	scalar and non-scalar value location (e.g.\ finding a scalar or record value in an array)
\item	differential local addressing (i.e.\ subscripting an array with an array)
\item	full or partial replication of shared read-only values
\item	full or partial replication of shared read-mostly values with automatic consistency management
\item	structures with rotating ownership, suitable for migration
\item	producer-consumer structures
\item	partitionable structures
\item	pre-determined run-time re-partitioning (i.e.\ re-distributing an array)
\item	dynamic re-partitioning (e.g.\ for load balancing)
\item	committed mutual exclusion (e.g.\ waiting on a semaphore)
\item	uncommitted mutual exclusion (e.g.\ lock or fail)
\item	barrier synchronization
\item	multiple concurrent barriers used by non-overlapping groups of processes
\item	fetch-and-add, and other fetch-and-operate functions
\item	pre-scheduled receipt of messages of known composition
\item	variable-length message receipt
\item	message selection by type
\item	message selection by source
\item	message selection by contents
\item	guarded message selection (e.g.\ CSP's {\tt{alt}} construct)
\item	broadcast and partial broadcast
\item	split-phase (non-blocking) operations
\item	data marshalling and unmarshalling of ``flat'' structures (e.g.\ arrays of scalars)
\item	data marshalling and unmarshalling of nested or linked structures
\item	heterogeneous parallelism (i.e.\ running different applications concurrently in one machine)
\item	pipelining
\item	distributed linked structures in systems without physically-shared memory
\item	indexing of distributed shared arrays
\item	collective I/O
\item	uniform-size interleaved I/O
\item	heterogeneous-size interleaved I/O
\item	independent I/O operations on a single shared file
\end{itemize}

\subsection{Memory Reference Patterns\label{s:cliche-memref}}

These are inspired by the categorization used in the Munin system \cite{b:munin-overview}.
Again, implementors are encouraged to phrase discussion of their work in these terms where appropriate.

>>>>

\section{Memory Reference Patterns\label{s:memory}}

These are inspired by the categorization originally used in the Munin system \cite{b:munin-overview}.
Again, implementors are encouraged to phrase discussion of their work in these terms where appropriate.

### 2009-08-31 paper/2009-09/method.tex
More work reformatting and reorganizing paper

The main aim of this work is to assess and compare
the ease with which programs can be developed and/or ported
using different parallel programming systems.
Here,
we discuss issues related to choosing the problems that make up the suite,
assessing the complexity of particular implementations,
and other purposes which implementations of these might serve.
Recognizing that these problems are extremely simple,
and easy to parallelize,
we refer to them as ``toys''.

@@

Our criteria for including problems in this suite are as follows:
\begin{enumerate}
\item	Each toy should require no more than an afternoon
	to write and test in a well-supported sequential language.
	We feel that making individual problems more complicated will only discourage implementation.
\item	The correctness of each implementation must be relatively easy to verify.
	Toys whose output is easily visualized are therefore to be preferred,
	as are toys whose results are insensitive to floating-point discretization effects.
\item	Each toy application should be amenable to
	an analysis of speedup or isoefficiency \cite{b:isoefficiency-tutorial}
	to allow a comparison of theory and practice.
\item	At least some toys should not be ``infinitely scalable''.
	Many real-world applications are not,
	and this suite should reflect such limitations.
\item	At least some toys should require I/O,
	since this important aspect of real-world programming is often neglected by PPS designers.
\item	There should be some potential redundancy in the type of parallelization done,
	so that re-use of software can be demonstrated.
\item	Together,
	the toys in the suite must exercise a wide range of common parallel operations
	(Appendix~\ref{s:cliche-ops})
	and memory reference patterns
	(Appendix~\ref{s:cliche-memref}).
\item	The ``obvious'' implementations of these toys should span
	a wide range of parallel programming paradigms,
	such as task farming,
	geometric and functional decomposition,
	and speculative parallelism.
\item	Parallelizations for several (if not all) of
	the types of programming systems shown in Figure~\ref{f:taxonomy} should be known.
	This suite is not intended to be a collection of research problems;
	reasonably simple and efficient implementations in most environments
	should be straightforward to produce.
\end{enumerate}

\begin{figure}
% \epsfysize=0.5\textheight
% \begin{center}\mbox{\epsffile{fig/taxonomy.eps}}\end{center}
\caption{A Taxonomy of Parallel Programming Systems\label{f:taxonomy}}
\end{figure}

\subsection{Specification\label{s:method-spec}}

Toys should be specified in terms of inputs and outputs,
rather than algorithmically.
For example,
``sort $N$ integers in the range $1{\ldots}R$,
where $R$ is not known in advance and there may be multiple instances of a given value''
is to be preferred over
``parallelize quicksort''.
As \cite{b:slalom} argued,
which parallel algorithm is best in a particular programming system
depends critically on the architectural assumptions made by that system.
Implementors should be free to employ those methods which they find natural and convenient.
The only exception we make to this rule is in the two matrix equation solvers,
where we require Gaussian elimination and successive over-relaxation.
Given the enormous number of matrix solution algorithms in use,
we felt we had to specify these particular algorithms in order to ensure that
different implementations of this suite would be comparable.

@@

These toys will exercise many different aspects of parallel systems.
However,
their ``single algorithm per program'' model is not representative of real applications.
These usually contain several discrete (and sometimes overlapping) phases,
each of which is qualitatively different.
A full implementation of this suite will therefore have two parts.
In the first,
each toy will be implemented as a stand-alone program.
In the second,
toys will be chained together to create the kind of phase behaviour seen in real applications
(Figure~\ref{f:chaining}).
This will test the ease with which heterogeneous parallelism can be mixed within a single program.
It will also show how well the system supports code re-use and information hiding,
which are crucial to the development of large programs.

@@

A second reason for chaining toys is that
many groups are now using algorithmic skeletons to structure parallel programs
\cite{b:cole-skel,b:p3l-overview,b:enterprise}.
Such skeletons encapsulate the details of a particular style of parallelism,
and hide low-level or platform-dependent issues.
We hope that chaining will show the strengths and weaknesses of such systems.

Finally,
chaining should be designed so that some toys can be executed concurrently.
Many PPSs impose extraneous constraints on programs,
e.g.\ require all processes to participate in every barrier synchronization,
or require the same executable to be loaded onto each processor.
These constraints can limit the exploitation of potential concurrency.
Permitting, but not requiring, concurrent execution of several toys should uncover such limitations.

@@

One crucial aspect of the specification of toys is
the way in which the sizes of problems are determined.
In a completely frozen model, the actual size of each problem would be compiled into each toy.
A fluid implementation, by contrast, would allow sizes to be specified at run-time,
and would only then allocate and partition data structures.
We have chosen to use an intermediate model (which might be called ``slushy''),
in which the maximum size of individual problems is specified during compilation,
but the actual size of a problem is only determined when the toy begins to execute.
We hope that the use of this model will force implementors to deal with
at least some of the software engineering issues
involved in building flexible large-scale applications,
without making implementation prohibitively difficult.

@@

There are at least two different ways in which usability might be compared.
One would be to measure the performance achieved by an ``average'' programmer
as a function of time on each of a range of problems.
This would lead to graphs of the kind shown in Figure~\ref{f:perf},
where the left graph shows a particular result
and the right graph a simplified abstraction of it.
By analogy with Hockney's $n_{1/2}$ measure \cite{b:hockney-perfparam},
which is the length of vector on which a pipelined architecture achieves
half its theoretical peak performance,
we could in principle find the value of $p_{1/2}$---the programming time required to achieve
half of a machine's peak performance\footnote{Note that
	if performance was measured as a fraction of the figures quoted by manufacturers for their machines,
	it is unlikely that the halfway mark would ever be reached.}
for a particular combination of programming system and problem type.

\begin{figure}
% \epsfxsize=12cm
% \begin{center}\mbox{\epsffile{fig/perf-time.eps}}\end{center}
\caption{Performance vs.\ Time\label{f:perf}}
\end{figure}

A second option would be to measure the complexities of
implementations of a single application using different programming systems.
With this approach,
comparative performance or speedup figures would be supplemented by complexity measures,
as shown in Figure~\ref{f:complex}.

\begin{figure}
% \epsfxsize=6cm
% \begin{center}\mbox{\epsffile{fig/perf-complex.eps}}\end{center}
\caption{Comparative Code Complexity\label{f:complex}}
\end{figure}

The drawback of the first approach is that there are no ``typical'' problems,
or ``average'' programmers.
Meaningful measurement of  $p_{1/2}$ would therefore require
measurement of a large number of programmers working on many problems.
Since this is too expensive to be practical,
the second approach is the one we have chosen to adopt.
It has drawbacks as well,
most notably the fact that no-one has ever demonstrated any better measure of program complexity
than a simple count of the number of lines in a program \cite{b:fenton-metrics}.
Since we anticipate a wide variation in the types of progamming systems that will be used in this work
(ranging from declarative dataflow languages
through functional languages with Lisp-like syntax
to parallel dialects of Fortran and C),
and in the indentation, commenting, and layout styles of particular programmers,
we may measure complexity by counting the number of tokens in the source code which are
keywords,
intrinsic operators,
user-defined functions,
type or variable declarations,
and miscellaneous punctuation.

@@

This suite is intended for assessing PPS usability,
but we envisage at least three other uses for implementations of it:
\begin{description}
\item[Related Research:]
	We hope that the existence of this suite will indicate
	to the developers of new parallel programming systems
	what a mature tool should be able to support.
	If sufficiently diverse,
	the suite could also be used to help test the correctness of new or improved systems.
	Finally,
	this suite could complement existing performance benchmarks,
	since performance figures for individual toys could be compared.
\item[Tool Development:]
	Requests from tool developers for existing parallel programs,
	or for event traces from them,
	are common on newsgroups such as {\tt{comp.parallel}}.
	We hope that traces from implementations of this suite
	will be used to compare different debugging and performance monitoring tools,
	that their source code will be used to test
	new parallelizing compilation techniques or mapping and scheduling algorithms,
	and so on.
\item[Teaching:]
	Implementations will be suitable for use as classroom examples,
	since they will be small enough to be understood quickly.
	The toys themselves should also be suitable as classroom exercises
	in a senior undergraduate course on parallel computing.
\end{description}

@@

Implementations of benchmark suites have usually been coded for absolute speed.
They are not representative of ``normal'' practice,
in which reasonable performance is the usual goal,
and so are inappropriate for this work.
In addition,
most existing benchmark suites are too large to be implemented or ported quickly
(e.g.\ SPEC~\cite{b:bench-over}),
over-specified (e.g.\ the Livermore Loops~\cite{b:livermore-loops}),
or concerned with a limited range of applications
(e.g.\ the NAS Parallel Benchmarks~\cite{b:nas-results}).
Finally,
most exist only in C or Fortran (or both);
many interesting parallel programming systems are built on top of other languages,
including Modula-2, Scheme, Prolog, and various dataflow languages.
We do not want the effort required to translate several large programs
to discourage researchers from participating in this exercise.

\vspace{\baselineskip}

@@

The short answer to this question is yes,
but no more than the results from performance benchmarks.
A longer answer is that	since we intend to measure the combination of code complexity and performance achieved,
rather than performance as a function of development time,
programmers will be allowed to revise and improve their implementations when and as desired.

>>>>

Our aim is to assess and compare the usability of parallel programming systems.
Here,
we discuss issues related to choosing the problems that make up our benchmark suite,
how to assess particular implementations,
and other purposes which implementations of these might serve.
Recognizing that our problems are extremely simple,
we refer to them as ``toys''.

@@

Our criteria for including toys are:

\begin{enumerate}

\item	Each toy should require no more than an hour
	to write and test in a well-supported sequential language.
	We feel that making individual problems more complicated will only discourage uptake.

\item	Correctness must be easy to verify.
	Toys whose output is easily visualized are therefore preferred,
	as are toys whose results are insensitive to floating-point effects.

\item	Speedup must similarly be easy to measure---while machine performance is not our focus,
	we are generally uninterested in parallel implementations that are slower than
	the sequential originals.

\item	At least some toys should not be ``infinitely scalable''.
	Many real-world applications are not,
	and this suite should reflect such limitations.

\item	At least some toys should require I/O,
	since this important aspect of real-world programming is often neglected by PPS designers.

\item	There should be some overlap in the toys' implementations,
	so that implementors can demonstrate how well
	their chosen systems take advantage of opportunities for code re-use.

\item	Together,
	the toys in the suite must exercise a wide range of parallel clich\'{e}s
	(Appendix~\ref{s:ops} and ~\ref{s:memory}).
	In particular,

\item	Toys should be specified by inputs and outputs rather than algorithmically,
	i.e.,
	``sort $N$ integers'' rather than ``parallelize quicksort'',
	so that implementors can choose algorithms that are ``natural'' for their systems.
	Implementations that parallelize a grossly inefficient sequential algorithm
	should expect to be criticized for doing so;
	see Section~\ref{s:toys-invperc} for more discussion of this.

\end{enumerate}

@@

The ``single algorithm per program'' model of many benchmarks is not representative of real programs,
which often contain several qualitatively different phases or modules.
A full implementation of this suite will therefore have two parts.
In the first,
each toy will be implemented as a stand-alone program.
In the second,
toys will be chained together as shown in Figure~\ref{f:chaining}.
This will test the ease with which heterogeneous parallelism can be mixed within a single program.
It will also show how well the system supports code re-use and information hiding,
which are crucial to the development of large programs.

@@

Where possible,
chaining should execute toys concurrently.
Some parallel programming systems impose extraneous constraints on programs,
e.g., require all processes to participate in every barrier synchronization,
or require the same executable to be loaded onto each processor.
These constraints can limit the exploitation of potential concurrency.
Permitting, but not requiring, concurrent execution of several toys should uncover such limitations.

@@

One crucial aspect of the specification of toys is
the way in which the sizes of problems are determined.
In a \emph{frozen} model,
the actual size of each problem and/or the number of processors available would be compiled into each toy.
A fully \emph{fluid} implementation, by contrast, would allow these sizes to be specified at run-time.
If a system does not support fully-fluid implementations,
it must at least use an intermediate model (which might be called \emph{slushy}),
in which the maximum size of individual problems is specified during compilation,
but the actual size of a problem is only determined when the toy begins to execute.

@@

There are at least two different ways in which usability could be compared.
One would be to measure the performance achieved by an ``average'' programmer
as a function of time on each toy.
By analogy with Hockney's $n_{1/2}$ measure \cite{b:hockney-perfparam},
we could in principle find the value of $p_{1/2}$---the programming time required to achieve
half of a machine's peak performance\footnote{Note that
	if performance was measured as a fraction of the figures quoted by manufacturers for their machines,
	it is unlikely that the halfway mark would ever be reached.}
for a particular combination of programming system and problem type.
However,
the number of implementors is likely to be too small for averages to be meaningful.

Alternatively,
we could use code metrics
to compare implementations of various toys.
This is also dubious,
as \cite{b:el-emam-metrics} has shown that code metrics do little more than measure lines of code.

We therefore intend to compare implementations through group consensus,
e.g.,
through interviews with implementors
and by asking other programmers who are familiar with the base languages used
to read and comment on the parallel versions of the toys.
We will also compare the length of code in parallel and sequential implementations,
though we realize that this can exaggerate the impact of parallelization.

@@

This suite is intended for assessing the usability of parallel programming systems,
but we envisage other uses.
First, this suite will indicate
what a mature parallel programming system should be able to support.
In particular,
we will ask implementors to describe how they debugged correctness and performance problems,
and in particular what tools they used.
(The lack of useful debugging tools is a chronic complaint among parallel programmers.)

This suite should also be useful for education,
since toys will be small enough to be understood quickly,
and parallelizing them should be a suitable classroom exercise
in a senior undergraduate course on parallel computing.

@@

Implementations of benchmark suites have usually been coded for absolute speed.
They are not representative of ``normal'' practice,
in which reasonable performance is the usual goal,
and so are inappropriate for this work.
In addition,
most existing benchmark suites are too large to be implemented or ported quickly
(e.g.\ SPEC~\cite{b:bench-over}),
over-specified (e.g.\ the Livermore Loops~\cite{b:livermore-loops}),
or concerned with a limited range of applications.
Finally,
most exist only in C or Fortran (or both);
many interesting parallel programming systems are built on top of other languages.
We do not want the effort required to translate several large programs
to discourage researchers from participating in this exercise.

@@

Yes,
but no more than the results from performance benchmarks.
A longer answer is that
since we intend to measure the combination of code complexity and performance achieved,
rather than performance as a function of development time,
programmers will be allowed to revise and improve their implementations when and as desired.
We hope this will encourage the evolutionary improvement seen in some online programming contests,
e.g., \cite{b:gulley-contest}.

### 2009-08-31 paper/2009-09/ops.tex
More work reformatting and reorganizing paper

\section{Parallel Clich\'{e}s\label{s:cliche}}

\subsection{Operations\label{s:cliche-ops}}

@@

\subsection{Memory Reference Patterns\label{s:cliche-memref}}

These are inspired by the categorization used in the Munin system \cite{b:munin-overview}.
Again, implementors are encouraged to phrase discussion of their work in these terms where appropriate.

\begin{itemize}
\item	Write-once:
	variables which are assigned a value once, and only read thereafter.
	These can be supported through replication.
\item	Write-many:
	variables which are written and read many times.
	If several processes write to the variable concurrently, they typically write to different portions of it.
\item	Producer-consumer:
	variables which are written to by one or more objects, and read by one or more other objects.
	Entries in a shared queue, or a mailbox used for communication between two processes, are examples.
\item	Private:
	variables which are potentially shared, but actually private.
	The interior points of a mesh in a program which uses geometric decomposition fall into this category,
	while boundary points belong to the previous class.
\item	Migratory:
	variables which are read and written many times in succession by a single process before ownership passes to another process.
	The structures representing particles in an $N$-body simulation are the clearest example of this category.
\item	Result:
	Accumulators which are updated many times by different processes, but thereafter read.
\item	Read-mostly:
	Any variable which is read much more often than it is written to.
	The best score so far in a search problem is a typical instance of this class:
	while any process might update it, in practice processes read its value much more often than they change it.
\item	Synchronization:
	Variables used to force explicit synchronization in a program, such as semaphores and locks.
	These are typically ignored for long periods, and then the subject of intense bursts of access activity.
\item	General read-write:
	Any variable which cannot be put in one of the above categories.
\end{itemize}

>>>>

\section{Parallel Operations\label{s:ops}}

@@

### 2009-08-31 paper/2009-09/prob-invperc.tex
More work reformatting and reorganizing paper

Invasion percolation models the displacement of one fluid
(such as oil)
by another
(such as water)
in fractured rock \cite{b:percolation-theory,b:invasion-percolation}.
In two dimensions,
this can be simulated by generating
an $N{\times}N$ grid of random numbers in the range $[1{\ldots}R]$,
and then marking the center cell of the grid as filled.
In each iteration,
one examines the four orthogonal neighbors of all filled cells,
chooses the one with the lowest value
(i.e.\ the one with the least resistance to filling),
and fills it in.
Figure~\ref{f:invperc1} shows the first few steps in this process,
while Figure~\ref{f:invperc2} shows the evolution of the final fractal shape.

>>>>

Invasion percolation models the displacement of one fluid
(such as oil)
by another
(such as water)
in fractured rock \cite{b:percolation-theory}.
In two dimensions,
this can be simulated by generating
an $N{\times}N$ grid of random numbers in the range $[1{\ldots}R]$,
and then marking the center cell of the grid as filled.
In each iteration,
one examines the four orthogonal neighbors of all filled cells,
chooses the one with the lowest value
(i.e.\ the one with the least resistance to filling),
and fills it in.
Figure~\ref{f:invperc1} shows the first few steps in this process,
while Figure~\ref{f:invperc2} shows the evolution of the final fractal shape.
